Fluent statistical computing interfaces for biological data analysis

### 2020-08-03 Rmd/05-embeddings.Rmd
update screenshots, strongly prompt the viewer to watch the videos

We *strongly* recommend viewing the linked videos for each case study
while reading. The videos presented show the visual 
appearance of the **liminal** interface, and how we can interact with the tour
via the controls previously described.

@@

We see in figure \@ref(fig:gaussian) and linked video that t-SNE has 
correctly split out each cluster and laid them out in a star like formation. 
This is agrees with the tour view,
where once we start the animation, the five clusters begin to appear but
generally points are more concentrated in the projection view compared to 
the t-SNE layout (figure \@ref(fig:gaussian)a). 
This can be confirmed via brushing the t-SNE view (figure \@ref(fig:gaussian)b).

@@

Next we view Gaussian clusters from the 
_Multi Challenge Dataset_, a benchmark simulation data set for clustering tasks
[@Rauber2009-vh]. This dataset has two Gaussian clusters with equal covariance 
embedded in 10-$d$, and a third cluster
with hierarchical structure. This cluster has two 3-$d$ clusters embedded in 
10-$d$, where the second cluster is subdivided into three smaller clusters, 
that are each equidistant from each other and have the same covariance structure.
Again t-SNE has correctly identified the sub-clusters but their relative 
locations to each other is distorted, with the orange and blue groups being
far from each other in the tour view (figure \@ref(fig:hierarchical)a). We see 
in this case that is difficult
to see the sub-clusters in the tour view, however, once we zoom and highlight
they become more apparent (figure \@ref(fig:hierarchical)b). When we brush 
the sub-clusters in the t-SNE, their
relative placement is again exaggerated, with the tour showing that they
are indeed much closer than the impression the t-SNE view gives.

@@

Next we explore some simulated noisy tree structured data (figure
\@ref(fig:fake-trees)), our interest here is how t-SNE visualisations break
topology of the data, and then seeing if we can resolve this by tweaking the
default parameters with reference to the global view of the data set. 
This simulation aims to mimick branching trajectories of cell differentiation:
if there were only mature cells, we would just see the tips of the branches 
which have a hierarchical pattern of clustering.

@@

We see that the t-SNE view has been unable to recapitulate the
topology of the tree - the backbone (blue) branch has been split into three
fragments (figure \@ref(fig:trees-01)a). We can see this immediately via the linked highlighting over both
plots. If we click on the legend for the zero branch, the blue coloured points
on each view are highlighted and the remaining points are made transparent.
From, here it becomes apparent from the tour view that the blue branch forms
the backbone of the tree and is connected to all other branches. We can also
again see that cluster sizes formed via t-SNE can be misleading; from the tour
view there is a lot of noise along the branches, while this does not appear to
be the case for the t-SNE result (figure \@ref(fig:trees-01)b).

From the first view, we modify the inputs to the t-SNE view, to try and produce
a better trade-off between local structure and retain the topology of the data.
We keep every parameter the same except that we initialise $Y$ with the first
two PCs (scaled to have standard deviation 1e-4) instead of the default random
initialisation and increase the perplexity from 30 to 100. We then combine
these results with our tour view as displayed in figure \@ref(fig:trees-02).

@@

To check and verify the clustering we construct a **liminal** view. 
We tour the first five PCs (approximately 20% of the variance in expression),
alongside the t-SNE view which was computed from all 25 PCs. 
Due to latency of the interface we do a weighted sample of  the rows based on
cluster membership, leaving us with approximately 10 per cent of the original
data size - 4,590 cells. Although this is not ideal, it still allows us to
get a sense of the shape of the clusters as seen in figure \@ref(fig:mouse-01).

@@

We observe that the embedding has mostly captured the
clusters relative location to each other to their location in high dimensional space, with a notable exception of points in cluster 3 and 10 as shown with
linked brushing (figure \@ref(fig:mouse-01)a). As expected, t-SNE mitigates the crowding problem that is
an issue for tour in this case, where points in clusters 2,4,6, and 11 are
clumped together in tour view, but are blown up in the embedding view (figure \@ref(fig:mouse-01)b). 
The tour appears to form a tetrahedron like shape, with points lying on
surface and along the vertices of the the tetrahedron in 5-$d$ PCA space -
a phenomena that has also been observed in @Korem2015-af 
(figure \@ref(fig:mouse-01)c). Brushing on
the tour view, reveals points in cluster 9 that are diffuse in the embedding 
view, points in cluster 9 are relatively far away and spread apart from other 
clusters in the tour view, but has points placed in cluster 3 and 9 in the embedding
(figure \@ref(fig:mouse-01)d).

Next, we identify marker genes for clusters using one sided Welch t-tests with
a minimum log fold change of one as recommended by @Amezquita2020-at, which
uses the testing framework from @McCarthy2009-qx. We select the top 10 marker 
genes that are upregulated in cluster 2, which was one of the clumped clusters
when we toured on principal components. Here the tour becomes an alternative
to a standard heatmap view for assessing shared markers; the basis
generation (shown as the biplot on the left view) reflects the relative 
weighting of each gene. We run the tour directly on the 
log normalised expression values using the same subset as before (figure \@ref(fig:mouse-02)).

@@

We see that the expression of the marker genes, appear to separate
the previously clumped clusters 2,4,6, and 11 from the other clusters, 
indicating that these genes are expressed in all four clusters (figure \@ref(fig:mouse-02)a). Using zooming, we can see a trajectory forming along
the clusters, while the axis view shows that magnitude of expression in 
the marker genes is similar across these separated clusters which is 
consistent with the results of marker gene analysis (figure \@ref(fig:mouse-02)b).

@@

```{r vimeo}
knitr::kable(dplyr::select(vid_lookup, `Case Study` = key, `URL` = url), 
             caption = "Case Study Videos",
             booktabs = TRUE)
```

>>>>

We *strongly* recommend viewing the linked videos for each case study
while reading. The videos are available to view online from the links
in the captions of each sections figures. The videos presented show the visual 
appearance of the **liminal** interface, and how we can interact with the tour
via the controls previously described. If you are unable to view the videos,
the figures in each case study consist of screenshots that summarise what is
learned from combining the tour and an embedding view.

@@

From the video linked inn figure \@ref(fig:gaussian), we learn that t-SNE has 
correctly split out each cluster and laid them out in a star like formation. 
This is agrees with the tour view,
where once we start the animation, the five clusters begin to appear but
generally points are more concentrated in the projection view compared to 
the t-SNE layout (figure \@ref(fig:gaussian)a). 
This can be seen via brushing the t-SNE view (figure \@ref(fig:gaussian)b).

@@

Next we view Gaussian clusters from the 
_Multi Challenge Dataset_, a benchmark simulation data set for clustering tasks
[@Rauber2009-vh]. This dataset has two Gaussian clusters with equal covariance 
embedded in 10-$d$, and a third cluster
with hierarchical structure. This cluster has two 3-$d$ clusters embedded in 
10-$d$, where the second cluster is subdivided into three smaller clusters, 
that are each equidistant from each other and have the same covariance structure.
From the video linked in figure \@ref(fig:hierarchical), we see 
that t-SNE has correctly identified the sub-clusters. However, their relative 
locations to each other is distorted, with the orange and blue groups being
far from each other in the tour view (figure \@ref(fig:hierarchical)a). We see 
in this case that is difficult
to see the sub-clusters in the tour view, however, once we zoom and highlight
they become more apparent (figure \@ref(fig:hierarchical)b). When we brush 
the sub-clusters in the t-SNE, their
relative placement is again exaggerated, with the tour showing that they
are indeed much closer than the impression the t-SNE view gives.

@@

Next we explore some simulated noisy tree structured data (figure
\@ref(fig:fake-trees)), our interest here is how t-SNE visualisations break
topology of the data, and then seeing if we can resolve this by tweaking the
default parameters with reference to the global view of the data set. 
This simulation aims to mimic branching trajectories of cell differentiation:
if there were only mature cells, we would just see the tips of the branches 
which have a hierarchical pattern of clustering.

@@

From the linked video, we see that the t-SNE view has been unable to 
recapitulate the topology of the tree - the backbone (blue) branch has been 
split into three fragments (figure \@ref(fig:trees-01)a). We can see 
this immediately via the linked highlighting over both
plots. If we click on the legend for the zero branch, the blue coloured points
on each view are highlighted and the remaining points are made transparent.
From, here it becomes apparent from the tour view that the blue branch forms
the backbone of the tree and is connected to all other branches. From the video 
it is easy to see that cluster sizes formed via t-SNE can be misleading; from the tour
view there is a lot of noise along the branches, while this does not appear to
be the case for the t-SNE result (figure \@ref(fig:trees-01)b).

From the first view, we modify the inputs to the t-SNE view, to try and produce
a better trade-off between local structure and retain the topology of the data.
We keep every parameter the same except that we initialise $Y$ with the first
two PCs (scaled to have standard deviation 1e-4) instead of the default random
initialisation and increase the perplexity from 30 to 100. We then combine
these results with our tour view as displayed in the linked video in the caption
of figure \@ref(fig:trees-02).

@@

To check and verify the clustering we construct a **liminal** view. 
We tour the first five PCs (approximately 20% of the variance in expression),
alongside the t-SNE view which was computed from all 25 PCs. 
Due to latency of the interface we do a weighted sample of  the rows based on
cluster membership, leaving us with approximately 10 per cent of the original
data size - 4,590 cells. Although this is not ideal, it still allows us to
get a sense of the shape of the clusters as seen from the linked video in figure \@ref(fig:mouse-01).

@@

From the video linked in figure \@ref(fig:mouse-01), we learn that the embedding has mostly captured the
clusters relative location to each other to their location in high dimensional space, with a notable exception of points in cluster 3 and 10 as shown with
linked brushing (figure \@ref(fig:mouse-01)a). As expected, t-SNE mitigates the crowding problem that is
an issue for tour in this case, where points in clusters 2,4,6, and 11 are
clumped together in tour view, but are blown up in the embedding view (figure \@ref(fig:mouse-01)b). 
The tour appears to form a tetrahedron like shape, with points lying on
surface and along the vertices of the the tetrahedron in 5-$d$ PCA space -
a phenomena that has also been observed in @Korem2015-af 
(figure \@ref(fig:mouse-01)c). Brushing on
the tour view, reveals points in cluster 9 that are diffuse in the embedding 
view, points in cluster 9 are relatively far away and spread apart from other 
clusters in the tour view, but has points placed in cluster 3 and 9 in the embedding
(figure \@ref(fig:mouse-01)d).

Next, we identify marker genes for clusters using one sided Welch t-tests with
a minimum log fold change of one as recommended by @Amezquita2020-at, which
uses the testing framework from @McCarthy2009-qx. We select the top 10 marker 
genes that are upregulated in cluster 2, which was one of the clumped clusters
when we toured on principal components. Here the tour becomes an alternative
to a standard heatmap view for assessing shared markers; the basis
generation (shown as the biplot on the left view) reflects the relative 
weighting of each gene. We run the tour directly on the 
log normalised expression values using the same subset as before.

@@

From the video linked in figure \@ref(fig:mouse-02), we see that the 
expression of the marker genes, appear to separate
the previously clumped clusters 2,4,6, and 11 from the other clusters, 
indicating that these genes are expressed in all four clusters (figure \@ref(fig:mouse-02)a). After zooming, we can see a trajectory forming along
the clusters, while the axis view shows that magnitude of expression in 
the marker genes is similar across these separated clusters which is 
consistent with the results of marker gene analysis (figure \@ref(fig:mouse-02)b).

@@

```{r vimeo}
vid_lookup$`Case Study` <- c(1, 2, 3,3,4,4)
knitr::kable(dplyr::select(vid_lookup, `Case Study`, 
                           `Example` = key, `URL` = url), 
             caption = "Case Study Videos",
             booktabs = TRUE)
```
